---
title: "Peer graded assignment"

date: "12/25/2021"
output: html_document
keep_md: true
---
## Title
Exploring the NOAA Storm Database: US Severe Weather impacts on Health and Economics
The goal of this assignment is to explore the NOAA Storm Database and answer questions about how severe weather events.The NOAA Storm Database is to be used to address the questions.The analysis consist of figures, and other summaries by using R to support the analysis.

## Synopsis
The NOAA database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. This shows how severe weather events can cause both public health and economic instabilities.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Processing
```{r}
storm <- read.table("repdata_data_StormData.csv", sep=",", header=TRUE)
Main_data <- storm
Main_data$BGN_DATE <- strptime(storm$BGN_DATE, "%m/%d/%Y %H:%M:%S")
Main_data <- subset(Main_data, BGN_DATE > "1995-12-31")
```
## Selecting Useful Data for this Analysis after study of NOAA Documentation by transforming data elements
```{r}
Main_data <- subset(Main_data, select = c(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
#cleaning event types names
Main_data$EVTYPE <- toupper(Main_data$EVTYPE)
# eliminating zero data
Main_data <- Main_data[Main_data$FATALITIES !=0 | 
                       Main_data$INJURIES !=0 | 
                       Main_data$PROPDMG !=0 | 
                       Main_data$CROPDMG !=0, ]
```
#Population health data processing
```{r}
Health_data <- aggregate(cbind(FATALITIES, INJURIES) ~ EVTYPE, data = Main_data, FUN=sum)
Health_data$PEOPLE_LOSS <- Health_data$FATALITIES + Health_data$INJURIES
Health_data <- Health_data[order(Health_data$PEOPLE_LOSS, decreasing = TRUE), ]
Top10_events_people <- Health_data[1:10,]
print(Top10_events_people)
```
#transforming letters and symbols to numbers for wconomic data processing
```{r}
Main_data$PROPDMGEXP <- gsub("[Hh]", "2", Main_data$PROPDMGEXP)
Main_data$PROPDMGEXP <- gsub("[Kk]", "3", Main_data$PROPDMGEXP)
Main_data$PROPDMGEXP <- gsub("[Mm]", "6", Main_data$PROPDMGEXP)
Main_data$PROPDMGEXP <- gsub("[Bb]", "9", Main_data$PROPDMGEXP)
Main_data$PROPDMGEXP <- gsub("\\+", "1", Main_data$PROPDMGEXP)
Main_data$PROPDMGEXP <- gsub("\\?|\\-|\\ ", "0",  Main_data$PROPDMGEXP)
Main_data$PROPDMGEXP <- as.numeric(Main_data$PROPDMGEXP)

Main_data$CROPDMGEXP <- gsub("[Hh]", "2", Main_data$CROPDMGEXP)
Main_data$CROPDMGEXP <- gsub("[Kk]", "3", Main_data$CROPDMGEXP)
Main_data$CROPDMGEXP <- gsub("[Mm]", "6", Main_data$CROPDMGEXP)
Main_data$CROPDMGEXP <- gsub("[Bb]", "9", Main_data$CROPDMGEXP)
Main_data$CROPDMGEXP <- gsub("\\+", "1", Main_data$CROPDMGEXP)
Main_data$CROPDMGEXP <- gsub("\\-|\\?|\\ ", "0", Main_data$CROPDMGEXP)
Main_data$CROPDMGEXP <- as.numeric(Main_data$CROPDMGEXP)

Main_data$PROPDMGEXP[is.na(Main_data$PROPDMGEXP)] <- 0
Main_data$CROPDMGEXP[is.na(Main_data$CROPDMGEXP)] <- 0
```
#creating total damage values
```{r}
library(dplyr)
Main_data <- mutate(Main_data, 
                    PROPDMGTOTAL = PROPDMG * (10 ^ PROPDMGEXP), 
                    CROPDMGTOTAL = CROPDMG * (10 ^ CROPDMGEXP))
#analyzing
Economic_data <- aggregate(cbind(PROPDMGTOTAL, CROPDMGTOTAL) ~ EVTYPE, data = Main_data, FUN=sum)
Economic_data$ECONOMIC_LOSS <- Economic_data$PROPDMGTOTAL + Economic_data$CROPDMGTOTAL
Economic_data <- Economic_data[order(Economic_data$ECONOMIC_LOSS, decreasing = TRUE), ]
Top10_events_economy <- Economic_data[1:10,]
print(Top10_events_economy)
```
## Results
graph one can concludes that TORNADOS, EXCESSIVE HEAT and FLOOD are the main contributors to deaths and injuries out of all event types of weather events.
```{r}
#plotting health loss
library(ggplot2)
g <- ggplot(data = Top10_events_people, aes(x = reorder(EVTYPE, PEOPLE_LOSS), y = PEOPLE_LOSS))
g <- g + geom_bar(stat = "identity", colour = "black")
g <- g + labs(title = "Total people loss in USA by weather events in 1996-2011")
g <- g + theme(plot.title = element_text(hjust = 0.5))
g <- g + labs(y = "Number of fatalities and injuries", x = "Event Type")
g <- g + coord_flip()
print(g)
```

graph two concludes that FLOOD, HURRICANE/TYPHOON and STORM SURGE are the main contributors to severe economic consequences out of all event types of weather events.
```{r}
#plotting economic loss
g <- ggplot(data = Top10_events_economy, aes(x = reorder(EVTYPE, ECONOMIC_LOSS), y = ECONOMIC_LOSS))
g <- g + geom_bar(stat = "identity", colour = "black")
g <- g + labs(title = "Total economic loss in USA by weather events in 1996-2011")
g <- g + theme(plot.title = element_text(hjust = 0.5))
g <- g + labs(y = "Size of property and crop loss", x = "Event Type")
g <- g + coord_flip()
print(g)
```

